\bfseries\large Phosphatidylinositol 4,5-bisphosphate regulates cilium transition zone maturation in \textit{Drosophila melanogaster}

### 2018-02-25 latex/paper.tex
Updates from edits

%% \usepackage{xcolor}
%% \pagecolor[RGB]{42,47,56}
%% \color[RGB]{189,189,161}

@@

\section{Introduction}
Cilia are sensory organelles present on almost all cells in the human body
\citep{satir2010primary}.
They function primarily by providing sensory inputs to cells, but can
also function as motile propellers in certain cells
\citep{bloodgood2010sensory}.
Consistent with their importance in cell signalling,
defects in cilium formation are associated with genetic disorders
known collectively as ciliopathies, which can display
neurological defects \citep{valente2014primary},
skeletal abnormalities \citep{hammarsjo2017novel, waters2011ciliopathies},
and infertility \citep{inaba2016sperm} in addition to other phenotypes.

A large number of ciliopathies are linked to mutations in proteins that
localize to the cilium transition zone (referred to herein as TZ).
The TZ is the proximal-most region of the cilium and
acts as a diffusion barrier that regulates the
bi-directional transport of protein and lipid cargo at the base of the cilium
\citep{reiter2012base, szymanska2012transition}.
The TZ is essential for cilium formation (ciliogenesis) and cilium-associated
signalling.
The conserved TZ protein CEP290, for example, is mutated in at least
six different ciliopathies, and is essential for cilium formation in both humans
and Drosophila \citep{basiri2014migrating}.
Although the protein composition of TZs has been investigated in various
studies ranging from single gene analysis to proteomics,
processes underlying TZ maturation are not as well-studied.

In the Drosophila male germline, nascent TZs first assemble on basal bodies
during G2 phase in spermatocytes and anchor to the plasma membrane.
This is followed by establishment of cilium-membrane connections
through a poorly understood mechanism
involving microtubule rearrangements \citep{gottardo2013cilium} and a
ciliary membrane that persists through meiosis \citep{riparbelli2012assembly}.
Ciliogenesis begins with the assembly of TZ proteins at the tips of
the nascent basal body.
This is followed by the anchoring of the TZ to the membrane and
establishment of a compartmentalized space bounded by the ciliary membrane
and the TZ.
TZ maturation has been described in
\textit{Paramecium} \citep{aubusson2015transition}
\textit{Caenorrhabditis elegans} \citep{serwas2017centrioles} and
\textit{Drosophila melanogaster} \citep{gottardo2013cilium}
and is most readily observed by an increase in TZ length from nascent to mature.
This are probably similar to the initial steps of TZ assembly and reorganization
in vertebrates.

Here, we show that the lipid phosphatidylinositol 4,5-bisphosphate (\PIP{})
is essential for proper TZ maturation in developing male germ cells in \textit{D. melanogaster}.
\PIP{} is a membrane lipid and one of seven phosphoinositides (PIPs)
present in eukaryotic cells.
It is primarily enriched at the plasma membrane and the base of cilia,
but its functions in ciliogenesis have not been studied.
We found that reduction of cellular \PIP{}
by expression of the bacterial PIP phosphatase SigD
or by inactivation of the PIP 5-kinase Skittles (Sktl) that localizes to cilia
induces longer than normal TZs without affecting TZ assembly.
These hyperelongated TZs show hallmarks of aberrant function similar to those
found in previously studied TZ mutants,
including loss of cilium-membrane associations.
We also find that the \textit{onion rings} allele of the
ciliopathy-associated \textit{EXOC8} gene encoding the human Exo84
decouples TZ hyperelongation from loss of cilium-membrane contacts.
Our results suggest that the plasma membrane lipid \PIP{} regulates TZ length,
maturation and function, acting to support ciliogenesis.

\section{Results}
\subsection{\PIP{} is essential for transition zone maturation}
To investigate whether reduction of cellular \PIP{} affects ciliogenesis in the
Drosophila male germline,
we used transgenic flies expressing the Salmonella phosphoinositide phosphatase SigD
under the $\beta$2-tubulin promoter (referred to herein as \sigd{}),
which is expressed exclusively in spermatocytes
\citep{wei2008depletion, fabian2010phosphatidylinositol}.
We previously showed that \sigd{} induces male sterility
due in part to defects in flagellar biosynthesis \citep{wei2008depletion}.
To understand whether this was due to aberrant TZ function,
we examined the localization of fluorescently-labelled
centriolar/basal body protein
Ana1 (CEP295 homolog) \citep{goshima2007genes, blachon2009proximal}
and the TZ protein Cep290 \citep{basiri2014migrating}
during early steps of cilium assembly in spermatocytes in both
wild-type controls and \sigd{}.
As shown in Figure 1A, Cep290 grew to much longer lengths in \sigd{}
compared to wild-type controls.
Ana1 lengths were not affected, and we did not observe an inverse
correlation between Cep290 and Ana1 length.
This phenotype was not specific to Cep290;
all TZ markers examined showed hyperelongation, including
acetylated tubulin \citep{gottardo2013cilium},
Chibby \citep{enjolras2012drosophila} and
MKS \citep{slaats2015mks1, vieillard2016transition} proteins (Figure 1).
TZ hyperelongation was a highly penetrant phenotype (\textgreater 90\%)
and showed high intra-cyst correlation (\textgreater 0.95) suggestive
of a dosage-based response to a shared cellular factor.
Hyperelongated TZs persisted through meiosis.
Despite TZ hyperlongation, however, axonemes were able to
elongate post-meiosis (Figure 1).
As we showed previously, the ultrastructure of these axonemes
was frequently aberrant,
either lacking symmetry completely or containing triplet microtubules instead of doublets
\citep{wei2008depletion}.

@@

\section{Discussion}
Cilia are essential for proper human development, and their disruption
can cause genetic disorders with severities ranging from congenital-
lethal \citep{}
to late onset \citep{}.
A large number of ciliopathies are associated with genes that encode for
proteins that function at the cilium transition zone (TZ).
Often, a single gene is linked to multiple ciliopathies with distinct phenotypes
and severities.
For example, \textit{CEP290} is linked to at least six different
cilium-associated human disorders.
It is likely that these mutations do not completely abolish CEP290
function, and may be affected by genetic modifiers in an affected individual
\citep{zaghloul2010functional}.
Animal models with partial loss of TZ function may therefore provide
insight on intermediate phenotypes associated with human disease.
Here, we found that reduced levels of the membrane lipid \PIP{},
which is known to be enriched at the base of cilia,
affects TZ maturation, but not assembly,
leading to hyperelongated TZs that show signs of functional defects, and
may represent one such modifier for TZ function.

@@

%%
%    FIGURE 1
%
\textbf{Figure 1. SigD expression induces transition zone hyperelongation.}
\begin{enumerate}[label={(\Alph*)}, nolistsep]
\item \sigd{} expression induces Cep290 hyperelongation in mature (late G2 phase) cilia.
\item Paired Ana1-Cep290 lengths in nascent and mature cilia ($n$ \textgreater 30).
\item \sigd{} expression induces hyperelongation of other TZ markers, such as Cby, Mks
  and acetylated tubulin.
\item TZ hyperelongation persists through meiosis but does not affect
  axoneme outgrowth.
\end{enumerate}

>>>>

@@

\section{Introduction}
Cilia are sensory organelles present on almost all cells in the human body
\citep{satir2010primary}.
They function primarily by providing sensory inputs to cells, but can
also function as motile propellers in certain cells
\citep{bloodgood2010sensory}.
Consistent with their importance in cell signalling,
defects in cilium formation are associated with genetic disorders
known collectively as ciliopathies, which can display
neurological defects \citep{valente2014primary},
skeletal abnormalities \citep{hammarsjo2017novel},
and infertility \citep{inaba2016sperm} in addition to other phenotypes
\citep{waters2011ciliopathies}.

A large number of ciliopathies are linked to mutations in proteins that
localize to the cilium transition zone (referred to herein as TZ).
The TZ is the proximal-most region of the cilium and
functions as a diffusion barrier that regulates the
bi-directional transport of protein and lipid cargo at the base of the cilium
\citep{reiter2012base, szymanska2012transition}.
The TZ is essential for cilium formation (ciliogenesis) and cilium-based
signalling.
The conserved TZ protein CEP290, for example, is mutated in at least
six different ciliopathies with a spectrum of clinical features and severities
, and is important for cilium formation
and function in both humans \citep{shimada2017vitro, stowe2012centriolar}
and Drosophila \citep{basiri2014migrating}.
Although the protein composition of TZs has been investigated in various
studies ranging from single gene analysis to proteomics,
processes underlying TZ maturation are not as well-studied.

Ciliogenesis begins with the assembly of TZ proteins at the tips of
the basal body.
This nascent TZ undergoes maturation, where the structure and protein
composition changes, allowing for anchoring of the TZ to the membrane and
establishment of a compartmentalized space bounded by the ciliary membrane
and the TZ.
In the Drosophila male germline, nascent TZs first assemble on basal bodies
during G2 phase in spermatocytes and anchor to the plasma membrane.
This is followed by establishment of cilium-membrane connections
through a poorly understood mechanism
involving microtubule rearrangements \citep{gottardo2013cilium}, and a
ciliary membrane that persists through meiosis \citep{riparbelli2012assembly}.
TZ maturation has been described in
\textit{Paramecium} \citep{aubusson2015transition}
\textit{Caenorrhabditis elegans} \citep{serwas2017centrioles} and
\textit{Drosophila melanogaster} \citep{gottardo2013cilium}
and is most readily observed by an increase in TZ length.
We expect that this process is analogous to the initial steps of TZ assembly
through sequential recruitment of TZ proteins and reorganization
in vertebrates.

Here, we show that the membrane lipid phosphatidylinositol 4,5-bisphosphate (\PIP{})
is essential for proper TZ maturation in developing male germ cells
in \textit{D. melanogaster}.
\PIP{} is relatively enriched within the plasma membrane and the base of cilia
in the male germline.
We show that expression of the bacterial PIP phosphatase SigD
or by inactivation of the PIP 5-kinase Skittles (Sktl) that localizes to cilia
induces longer than normal TZs without affecting TZ assembly.
These hyperelongated TZs show hallmarks of aberrant function similar to those
found in previously studied TZ mutants,
including loss of cilium-membrane associations.
We also find that the \textit{onion rings} allele of the
ciliopathy-associated \textit{EXOC8} gene encoding the human Exo84
decouples TZ hyperelongation from loss of cilium-membrane contacts.
Our results suggest that the plasma membrane lipid \PIP{} regulates TZ length,
maturation and function, thereby acting to support ciliogenesis.

\section{Results}
\subsection{\PIP{} is essential for transition zone maturation}
To investigate whether reduction of cellular \PIP{} affects ciliogenesis in the
Drosophila male germline,
we used transgenic flies expressing the Salmonella phosphoinositide phosphatase SigD
under the $\beta$2-tubulin promoter (referred to herein as \sigd{}),
which is expressed exclusively in spermatocytes
\citep{wei2008depletion, fabian2010phosphatidylinositol}.
We previously showed that \sigd{} induces male sterility
due in part to defects in flagellar biosynthesis \citep{wei2008depletion}.
To understand whether this was due to aberrant TZ function,
we examined the localization of fluorescently-labelled
centriolar/basal body protein
Ana1 (CEP295 homolog) \citep{goshima2007genes, blachon2009proximal}
and the TZ protein Cep290 \citep{basiri2014migrating}
during early steps of cilium assembly in spermatocytes in both
wild-type controls and \sigd{}.
As shown in Figure 1A,
while Cep290 on nascent TZs look similar in wild-type and \sigd{}, 
it grew to significantly longer lengths in \sigd{}
compared to wild-type controls.
Ana1 lengths were not affected, and we did not observe an inverse
correlation between Cep290 and Ana1 length.
This hyperelongation phenotype was not specific to Cep290;
all TZ markers examined showed hyperelongation, including
acetylated tubulin \citep{gottardo2013cilium},
Chibby \citep{enjolras2012drosophila} and
MKS \citep{slaats2015mks1, vieillard2016transition} proteins (Figure 1).
TZ hyperelongation was a highly penetrant phenotype (\textgreater 90\%)
and showed high intra-cyst correlation (\textgreater 0.95) suggestive
of a dosage-based response to a shared cellular factor,
presumably SigD.
Hyperelongated TZs persisted through meiosis.
Despite TZ hyperlongation, however, axonemes were able to
elongate post-meiosis (Figure 1).
As we showed previously, the ultrastructure of these axonemes
was frequently aberrant,
either lacking symmetry completely or containing triplet microtubules
instead of doublets \citep{wei2008depletion}.

@@

\section{Discussion}
Cilia are essential for proper human development, and their disruption
can cause genetic disorders with severities ranging from congenital-
lethal \citep{}
to late onset \citep{}.
A large number of ciliopathies are associated with genes that encode for
proteins that function at the cilium transition zone (TZ).
Often, a single gene is linked to multiple ciliopathies with
a spectrum of phenotypes and severities.
For example, \textit{CEP290} is linked to at least six different
cilium-associated human disorders.
It is likely that these mutations do not completely abolish CEP290
function, and may be affected by genetic modifiers in an affected individual
\citep{zaghloul2010functional, khanna2009common, leitch2008hypomorphic}.
Animal models with partial loss of TZ function may therefore provide
insight on intermediate phenotypes associated with human disease.
Here, we found that reduced levels of the membrane lipid \PIP{},
which is known to be enriched at the base of cilia,
affects TZ maturation, but not assembly,
leading to hyperelongated TZs that show signs of functional defects, and
may represent one such modifier for TZ function.

@@

%%
%    FIGURE 1
%
\textbf{Figure 1. SigD expression induces transition zone hyperelongation.}
\begin{enumerate}[label={(\Alph*)}, nolistsep]
\item \sigd{} expression induces Cep290 hyperelongation in mature cilia.
  Stages correspond to descriptions in \citep{cenci1994chromatin}.
\item Paired Ana1-Cep290 lengths in nascent and mature cilia ($n$ \textgreater 30).
\item \sigd{} expression induces hyperelongation of other TZ markers, such as Cby, Mks
  and acetylated tubulin.
\item TZ hyperelongation persists through meiosis but does not affect
  axoneme outgrowth.
\end{enumerate}
